Death By Dropdown?

### 2025-04-07 conference-materials/2025_shiny-conf/slides.qmd
Enhances presentation with improved speaker notes
- Refines speaker notes for clarity and engagement, emphasizing the role of UI/UX in Shiny development. - Adds notes about simplified examples to aid understanding.
- Updates psychological concepts on slide 11 to better align with presentation goals

::: {.notes}
**Opening Hook**: Begin with energy! Make eye contact across the room.

- "Raise your hand if you've ever encountered a dashboard that looked like this..." (gesture to image)
- "Dashboard rot" = dashboards built but never used - costing real development time and money
- Ask: "How many of you have abandoned using a dashboard because it was too complex?" (engage audience)
:::

@@

- Key message: "The answer isn't more data - it's better storytelling with data"
- "SoundPulse, a music streaming analytics team, faced this exact problem"
  * They're a composite client
- Their market research team needed to present competitor analysis to executives
  * Typically involves turning tons of complex analyses into non-technical insights
- "We're going to show you exactly how psychological principles transformed their dashboard - with approaches and code you can use immediately"
:::

@@

- Even though SoundPulse is a composite client...this is representative of real dashboards we've seen across the market research industry
- Point out key issues:
  * "Filter overload before seeing ANY insights"
  * "Multiple tabs with no clear starting point"
  * "Complete cognitive overload"
- "This is why dashboards get abandoned - nobody wants this experience"
:::

@@

::: {.notes}
**Framework Introduction**: Slow down here - this is the core methodology.

- "The 5 stages form an easy workflow: Notice → Interpret → Structure → Anticipate → Validate"
  * Click through each
- "Each stage translates psychological principles into practical developer actions"
- "This isn't just theory - it's a step-by-step process you can implement today"
- "These stages are iterative - you'll often revisit earlier stages as you discover new insights"
  * Any order
:::

@@

- **Stage 1 (Notice):** The first thing was users were overwhelmed with options
  * This is what we call 'cognitive load' - when the brain gets overtaxed with information
  * We was a violation of Hick's Law, which simply means that more choices lead to longer decision times and confusion
- **Stage 2 (Interpret):** We then focused on something called 'Processing Fluency' - making the information easier to digest
  * For example, using color to create a visual hierarchy of information so the brain can process it more easily
- **Stage 3 (Structure):** "We then restructured the UI using the 'Principle of Proximity' - which is a fancy way of saying we started grouping related elements together. 
  * Now there are a ton of ways you can do this...
  * But as you'll see in our code example, we created groups of primary and secondary metrics to create a natural flow through the data
- Now we'll show you the actual code implementation of these principles
:::

@@

- For each image as it appears:
  * First image: "Notice how grouping reduces cognitive load"
  * Second image: "The consistent legend simplifies comprehension"
  * Third image: "Color coding creates instant visual processing"
  * Fourth image: "Proximity principles group related metrics"
- "This is the power of intentional design - every element has a psychological purpose"
:::

@@

- "This first example shows how just a few lines implement powerful BID principles"
- "groupBy reduces cognitive load while creating visual hierarchy"
- "Color coding isn't just aesthetic - it creates instant processing fluency"
- "Progressive disclosure means users only see details when needed"
- "These three techniques alone dramatically improve dashboard usability"
- "This simple implementation already reduces cognitive load and enhances processing - but we can do more"
:::

@@

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "4-14|15-49"

# Packages: reactable, dplyr, bslib
reactable(
  soundpulse_data,
  # Stage 3: Group related elements (Principle of Proximity)
  columnGroups = list(
    colGroup(
      name = "Core Metrics",
      columns = c("Metric 1", "Metric 2", "Metric 3", "Metric 4")
    ),
    colGroup(
      name = "Key Performance",
      columns = c("KPI 1", "KPI 2", "KPI 3")
    )
  ),
  # Stage 1: Dynamic legend reduces choice complexity (Hick's Law)
  defaultColDef = colDef(
    header = function(value) {
      layout_inline(
        value,
        span(
          style = css(
            display = "inline-block",
            margin_left = "5px",
            height = "10px", 
            width = "10px",
            border_radius = "50%",
            background_color = "#1c6d7d"
          )
        )
      )
    }
  ),
  # Combine multiple visual elements for age groups
  columns = list(
    Age = colDef(
      cell = function(value) {
        colors <- c(
          "18-24" = "#1c6d7d",
          "25-34" = "#35a4ae", 
          "35-44" = "#98d3ca",
          "45-54" = "#ffca90",
          "55-64" = "#1c6d7d",
          "65-74" = "#dd8500"
          )

@@

- "columnGroups implements Principle of Proximity - grouping primary and secondary metrics together"
- "The dynamic legend reduces choice complexity - addressing Hick's Law"
- "Color coding by demographic creates instant pattern recognition"
- "Notice how each feature directly maps to a BID principle"
- Ask: "How many have used columnGroups in reactable before?" (engagement opportunity)
  * If not many...great, we taught something!
  * If a lot...could also mention (quickly) how to extend with legend/div
:::

@@

::: {.column width="48%"}
- <span style="color:#ffca90">**Stage 4: Anticipate**</span>
  * Executives comparing metrics without context (*Anchoring Effect*)
  * Teams interpreting data differently (*Confirmation Bias*)
  * Different perspectives needed for different teams (*Framing & Loss Aversion*)
:::

::: {.column width="48%"}
- <span style="color:#98d3ca">**Stage 5: Validate**</span>
  * Teams needed actionable summaries (*Peak-End Rule*)
  * No clear path to consensus<br>(*Cooperation & Coordination*)
:::
::::

@@

- "For the later stages of the BID framework, we addressed more subtle psychological factors that affect how users interpret and act on data."
- **Stage 4 (Anticipate):** We noticed executives were making flawed comparisons without proper context.
  * This relates to the 'Anchoring Effect' - where people rely too heavily on the first piece of information they see
  * We also saw teams struggling to interpret the data in the same way, so we addressed this with something called a "framing toggle"...more on that in just a moment
- **Stage 5 (Validate):** Finally, we applied the 'Peak-End Rule' - the idea that people judge experiences mainly by their most intense point and how they end
  * We leveraged this by creating summary cards that highlight key takeaways, increasing the chances that users leave the app with a clear, memorable impression of the most important insights
:::

@@

- For each image as it appears:
  * First image: "Gauge charts provide instant comprehension"
  * Second image: "The framing toggle changes perspective while maintaining data integrity"
  * Third image: ""Summary cards implement the Peak-End Rule"
- "Also, the card-based layout creates clear visual hierarchy and reduces cognitive load"
- "Every element in this dashboard has a specific psychological purpose"
- "This is what we mean by behavior-informed design - intentional choices based on how humans think"
:::

@@

- This directly addresses the Anchoring Effect from Stage 4"
- "Without reference lines, metrics like '65%' lack meaning - is it good or bad?"
- "Benchmarks provide the context users need for proper interpretation...and helps with alignment across users"
- "The distinctive orange dashed line creates visual separation from the data"
- "This simple addition dramatically improves decision quality"
:::

@@

- "This implementation shows how framing affects perception - same data, different decisions"
- "The 'Progress' frame emphasizes achievement - how far we've come"
- "The 'Gap' frame emphasizes what remains - how far we still need to go"
- "Different teams at SoundPulse used different frames - marketing used Progress, product used Gap"
- "The server-side code dynamically recalculates values based on the framing perspective"
- "This is psychological design directly impacting business decisions"
:::

@@

- This is a great example to illustrate the theory behind these concepts
- "In this case we are showing the *Same exact data*...but users experience completely different psychological impact"
- "Progress framing creates motivation and highlights achievement"
  - "Marketing preferred progress framing to build stakeholder confidence"
- "Gap framing creates urgency and focuses on what remains to be done"
  - "Product teams used gap framing to prioritize development"
- "This isn't just theory - a classic JDM study presented a 90% survival rate vs. a 10% mortality rate - people react completely differently"
- "This is why understanding psychological principles is so powerful for dashboard design"
- "Think about your own dashboards - are you framing data in a way that drives the right decisions?"
:::

@@

- "We're not just sharing a framework or theories - we're giving you tools to implement it immediately"
- "The `{bidux}` package makes these principles accessible to every R developer"
- "Phase 1 is available today on GitHub"
- Future phases will include LLM integration and ready-to-use UI components"
- "Think of `{bidux}` as your behavioral science assistant for Shiny app development"
:::

@@

- "Let's look at what {bidux} offers today"
- "The concept dictionary provides implementation advice, not just theory"
- "You can search across psychology, UI/UX, and data visualization fields...and more!"
- "The workflow functions guide you through each BID stage"
- "Auto-suggestion of theories based on your problem description"
- "Component suggestions provide concrete UI implementation ideas"
- When the link appears: "We're launching this at ShinyConf today - check out the repo for full documentation"
:::

@@

- "The results speak for themselves"
- "Users who previously abandoned the dashboard now rely on it daily"
- "Decisions that took days now happen in minutes"
- We're currently building telemetry into our dashboards to gather more precise measurements...but the message from our users is already clear: this framework is changing how teams interact with their data
- When testimonial appears: "This transformation is what BID delivers"
- "Remember, this wasn't just about visual improvements - this was a fundamental shift in how our users...and developers...approached the dashboard"
:::

@@

- "BID reduces friction between users and insights"
- "The tools we've shown today give you everything you need to get started"
- "`{bidux}` is available now to help structure your implementation"
- "Every dropdown you eliminate helps someone make a better decision"
- "This is about making your users successful, not just making dashboards"
:::

@@

- We'd love to hear about your dashboard challenges"
- "Don't forget to follow us on LinkedIn/GitHub for updates on {bidux} development"
- "Feel free to connect with us for any questions or deeper discussions!"
- Quick answers to likely questions:
  * "How do you measure success?" → "Start with user testing and session analytics"
  * "Does this work for technical audiences?" → "Absolutely, technical users appreciate clarity too"
  * "How long does implementation take?" → "Start small - even one principle makes a difference"
:::

>>>>

::: {.notes}
**Opening Hook**: Begin with energy!

- "Raise your hand if you've ever encountered a dashboard that looked like this..." (gesture to image)
- "Dashboard rot" = dashboards built but never used - costing real development time and money
- Ask: "How many of you have abandoned using a dashboard because it was too complex?" (engage audience)
- "UI/UX is often the most overlooked part of Shiny development, yet it's the part users interact with most directly"
:::

@@

- Key message: "The answer isn't more data - it's better storytelling with data"
- "SoundPulse, a music streaming analytics team, faced this exact problem"
  * They're a composite client based on our experiences with multiple market research clients
- Their market research team needed to present competitor analysis to executives
  * Typically involves turning tons of complex analyses into non-technical insights
- "We're going to show you exactly how psychological principles transformed their dashboard - with approaches and code you can use immediately"
:::

@@

- Even though SoundPulse is a composite client...this is representative of many dashboards we've seen across the market research industry
- Point out key issues:
  * "Filter overload before seeing ANY insights"
  * "Multiple tabs with no clear starting point"
  * "Complete cognitive overload"
- "This is why dashboards get abandoned - nobody wants this experience"
- "Throughout the presentation, we'll show you the transformation process and the principles that guided our decisions"
:::

@@

::: {.notes}
**Framework Introduction**: Slow slightly to introduce - but drive home in next slide.

- "The 5 stages form an easy workflow: Notice → Interpret → Structure → Anticipate → Validate"
  * Click through
- "Notice is about identifying objective friction points - where users struggle with the interface"
- "Interpret goes deeper to understand what story or insight users truly need - moving from symptoms to diagnosis"
- "Structure focuses on organizing information in ways that align with how our brains process information"
- "Anticipate considers how cognitive biases might affect data interpretation"
- "Validate ensures users leave with clear takeaways and a positive final impression"
- "These stages are iterative - you'll often revisit earlier stages as you discover new insights"
- "This diagram will serve as our roadmap throughout the presentation"
:::

@@

- **Stage 1 (Notice):** Users were overwhelmed with options
  * This is related to 'cognitive load' - which is the mental effort needed to process information
  * When the brain gets overtaxed, comprehension drops
  * Hick's Law also shows decision time increases logarithmically with the number of choices 
  * So more options, means exponentially more confusion
- **Stage 2 (Interpret):** Taking this to the next step helped us interpret what the user needed: Increased Processing Fluency
  * Processing fluency is the ease with which information is processed - when something is easy to process, people find it more credible and trustworthy
  * We also identified the key story the executives needed to understand...rather than showing all possible data
- **Stage 3 (Structure):** Moving to the Structuring stage...applied the 'Principle of Proximity' - which is a fancy way of saying we started grouping related elements together. 
  * Dual-Processing Theory distinguishes between fast, intuitive thinking and slower, deliberative thinking - we designed for both
  * We created groups of primary and secondary metrics to create a natural flow through the data
- "Now we'll show you the actual code implementation of these principles"
:::

@@

- For each image as it appears:
  * First image: "Notice how grouping reduces cognitive load"
  * Second image: "The consistent legend simplifies comprehension"
  * Third image: "Color coding creates instant visual processing"
  * Fourth image: "Proximity principles group related metrics"
- "This is the power of intentional design - every element has a psychological purpose"
- "These examples demonstrate the end result we achieved with SoundPulse. The code examples coming up show the core concepts we used, simplified for presentation clarity."
:::

@@

- "These examples demonstrate core concepts that translate to what we implemented with SoundPulse. While simplified for clarity, they use the same packages and techniques."
- "groupBy reduces cognitive load while creating visual hierarchy"
- "Color coding isn't just aesthetic - it creates instant processing fluency"
- "Progressive disclosure means users only see details when needed"
- "These three techniques alone dramatically improve dashboard usability"
- "When implementing these techniques, consider your data volume and complexity. For larger applications, you might need to balance visual richness with performance."
- "For implementation specifics beyond what we can cover today, feel free to connect with us afterward."
:::

@@

```{r}
#| eval: false
#| echo: true
#| code-line-numbers: "4-14|15-50"

# Packages: reactable, dplyr, bslib
reactable(
  soundpulse_data,
  # Stage 3: Group related elements (Principle of Proximity)
  columnGroups = list(
    colGroup(
      name = "Core Metrics",
      columns = c("Metric 1", "Metric 2", "Metric 3", "Metric 4")
    ),
    colGroup(
      name = "Key Performance",
      columns = c("KPI 1", "KPI 2", "KPI 3")
    )
  ),
  # Stage 1: Visual indicators for headers 
  # (simplification of Hick's Law dynamic legend)
  defaultColDef = colDef(
    header = function(value) {
      layout_inline(
        value,
        span(
          style = css(
            display = "inline-block",
            margin_left = "5px",
            height = "10px", 
            width = "10px",
            border_radius = "50%",
            background_color = "#1c6d7d"
          )
        )
      )
    }
  ),
  # Combine multiple visual elements for age groups
  columns = list(
    Age = colDef(
      cell = function(value) {
        colors <- c(
          "18-24" = "#1c6d7d",
          "25-34" = "#35a4ae", 
          "35-44" = "#98d3ca",
          "45-54" = "#ffca90",
          "55-64" = "#1c6d7d",
          "65-74" = "#dd8500"
          )

@@

- "columnGroups implements Principle of Proximity - grouping primary and secondary metrics together"
- "The dynamic legend reduces choice complexity - addressing Hick's Law"
- "Color coding by demographic creates instant pattern recognition"
- "Notice how each feature directly maps to a BID principle - this is intentional design, not just aesthetics"
- "Start with simpler implementations and add complexity as needed. We've layered these features progressively."
- Ask: "How many have used columnGroups in reactable before?" (engagement opportunity)
  * If not many...great, we taught something!
  * If a lot...could also mention (quickly) how to extend with legend/div
:::

@@

::: {.column width="48%"}
- <span style="color:#ffca90">**Stage 4: Anticipate**</span>
  * Executives comparing metrics without context (*Anchoring Effect*)
  * Different perspectives needed for different teams (*Framing & Loss Aversion*)
:::

::: {.column width="48%"}
- <span style="color:#98d3ca">**Stage 5: Validate**</span>
  * Teams needed actionable summaries (*Peak-End Rule*)
:::
::::

@@

- "For the later stages of the BID framework, we addressed more subtle psychological factors that affect how users interpret and act on data."
- **Stage 4 (Anticipate):** We noticed executives were making flawed comparisons without proper context.
  * This relates to the 'Anchoring Effect' - where people rely too heavily on the first piece of information they see
  * We also saw teams struggling to interpret the data in the same way, so we addressed this with something called a "framing toggle"...more on that in just a moment
- **Stage 5 (Validate):** Finally, we applied the 'Peak-End Rule' - the idea that people judge experiences mainly by their most intense moment and ending
  * We created summary cards highlighting key takeaways so users leave with a clear, memorable impression
:::

@@

- For each image, as it appears:
  * First image: "Gauge charts provide instant comprehension through visual metaphor"
  * Second image: "To help with the 'Anchoring Effect', we added reference lines to provide more accurate anchoring points"
  * Third image: "The framing toggle changes perspective while maintaining data integrity"
  * Fourth image: "Summary cards implement the Peak-End Rule, ensuring users leave with key insights"
- "Also, the card-based layout creates clear visual hierarchy and reduces cognitive load"
- "Every element in this dashboard has a specific psychological purpose"
- "This is what we mean by behavior-informed design - intentional choices based on how humans think"
- "These visuals represent our final implementation with SoundPulse, which we'll now break down into code components"
:::

@@

- "This directly addresses the Anchoring Effect from Stage 4"
- "The Anchoring Effect occurs when people rely too heavily on the first piece of information encountered - in data visualization, this often means making judgments without proper context"
- "Without reference lines, metrics like '65%' lack meaning - is it good or bad?"
- "Benchmarks provide the context users need for proper interpretation...and helps with alignment across users"
- "The distinctive orange dashed line creates visual separation from the data"
- "This simple addition dramatically improves decision quality by providing a meaningful anchor point"
:::

@@

- "This implementation shows how framing affects perception - same data, different decisions"
- "The code shows both UI and server components working together to implement framing"
- "While this is a simplified example, it demonstrates the core concept we implemented in the SoundPulse dashboard"
- "The 'Progress' frame emphasizes achievement - how far we've come"
- "The 'Gap' frame emphasizes what remains - how far we still need to go"
- "The server-side code dynamically recalculates values based on the framing perspective"
- "This is psychological design directly impacting business decisions"
:::

@@

- "This is a great example to illustrate the theory behind these concepts"
- "In this case we are showing the *Same exact data*...but users experience completely different psychological state"
- "Progress framing creates motivation and highlights achievement"
  - "Marketing preferred progress framing to build stakeholder confidence"
- "Gap framing creates urgency and focuses on what remains to be done"
  - "Product teams used gap framing to prioritize development"
- "This isn't just theory - a classic study in the Judgment and Decision Making literature presented identical medical treatment outcomes as either a 90% survival rate or a 10% mortality rate - people react completely differently to these framings"
- "This is why understanding psychological principles is so powerful for dashboard design"
- "Think about your own dashboards - are you framing data in a way that drives the right decisions?"
:::

@@

- "We're not just sharing a framework or theories - we're giving you tools to implement it immediately"
- "The `{bidux}` package directly implements the 5-stage BID framework we've discussed today"
- "Each function maps to a specific stage and guides your thinking through that part of the development process"
- "Phase 1 is available today on GitHub - it provides the conceptual foundation and workflow guidance"
- "Future phases will include LLM integration to suggest specific implementations based on your needs"
- "We're also developing a UI component library with pre-built elements that embody these principles"
- "Think of `{bidux}` as your behavioral science assistant for Shiny app development"
:::

@@

- "Let's look at what {bidux} offers today"
- "The concept dictionary provides implementation advice, not just theory"
- "You can search across psychology, UI/UX, and data visualization fields...and more!"
- "The workflow functions guide you through each BID stage"
- "Auto-suggestion of theories based on your problem description"
- "Component suggestions provide concrete UI implementation ideas"
- "The package helps connect psychological principles to specific suggestions and components"
- "While we don't have time to go through a complete workflow demonstration, the package documentation on GitHub provides detailed examples"
- When the link appears: "We're launching this at ShinyConf today - check out the repo for full documentation"
:::

@@

- "The results speak for themselves"
- "While our quantitative metrics are still being collected through new telemetry systems we're implementing, qualitative feedback has been overwhelmingly positive"
- "Users who previously abandoned the dashboard now rely on it daily"
- "Decisions that took days now happen in minutes"
- It's still early, but the emerging message from our clients is clear: this framework is changing how teams interact with their data
- When testimonial appears: "This transformation is what BID delivers"
- "Remember, this wasn't just about visual improvements - this was a fundamental shift in how our users...and developers...approached the dashboard"
:::

@@

- "BID reduces friction between users and insights by applying psychological principles in a systematic way"
- "The tools we've shown today - reactable and echarts4r - give you everything you need to get started"
- "The bidux package makes these principles accessible to every R developer, guiding you through each stage"
- "Start with small implementations of these principles - even one or two can dramatically improve your dashboard"
- "Every dropdown you eliminate or visual cue you add helps someone make a better decision"
- "This is about making your users successful, not just making dashboards"
- "UI/UX is often the most overlooked part of Shiny development, yet it's where behavioral science can have the greatest impact"
:::

@@

- We'd love to hear about your dashboard challenges"
- "Don't forget to follow us on LinkedIn/GitHub for updates on {bidux} development"
- "Feel free to connect with us for any questions or deeper discussions about implementation details we couldn't cover today"
- Quick answers to likely questions:
  * "How do you measure success?" → "Start with user testing and session analytics; we're building telemetry tools to help quantify impact"
  * "Does this work for technical audiences?" → "Absolutely, technical users appreciate clarity too - in fact, they often have the highest cognitive load due to the complexity of their work"
  * "How long does implementation take?" → "Start small - even one principle makes a difference. A complete BID implementation can take 1-2 sprints depending on dashboard complexity"
  * "How does this apply to domains beyond music?" → "We've successfully applied these principles in healthcare, finance, and retail dashboards - the psychological principles are universal"
:::
